---
title             : "Residential Instability and Childhood Wellbeing"
shorttitle        : "Residence and Wellbeing"

author: 
  - name          : "Libby Kramer"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "75 E River Parkway, Minneapolis MN"
    email         : "krame937@umn.edu"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"

affiliation:
  - id            : "1"
    institution   : "University of Minnesota"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  Research has suggested that quality of housing is a social determinant of health and that residential stability is a factor for older adults's health (CITE). This report will look more specifically at the relationship between residential stability and child well-being. As housing is not gaurenteed in the United States it is important to understand the relationship between where children are living and their wellbeing at home and in school. 
  
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  One or two sentences to put the results into a more **general context**.
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "housing, residential stability, childhood wellbeing, health, social determinants of health"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library("groundhog")
library("dplyr")
library("psych")
library("labelled")
library("ggplot2")
library("data.table")
library("dplyr")
library("bit64")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(0628)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

```{r data setup, include = FALSE}
# selecting variables from the full SIPP data set 
ds <- c("pu2024 2.csv")
pu <- fread(ds, sep = "|", select = c("ECLUB", "EEXPSCH", "EREPGRD", "ESPORT", "EEHC_PVTEN", "EEHC_WHY", "TAGE", "TDOB_BYEAR", "ERACE", "ESEX"))

names(pu) <- toupper(names(pu))

head(pu, 20)

# creating objects to be inserted into the report 
# omitting missing values from the data set
dat <- na.omit(pu)

# assigning variables new names
cwb_1 <- (dat$ECLUB)

# key cwb_2 negatively
cwb_2 <- (dat$EEXPSCH)
recode(dat$EEXPSCH, '1'= 2, '2' = 1)
cwb_2_r <- recode(cwb_2, '1'= 2, '2' = 1)
cwb_2_r

# checks that cwb_2 was reverse coded correctly
table(cwb_2)
table(cwb_2_r)

# key cwb_3 negatively
cwb_3 <- (dat$EREPGRD)
recode(dat$EREPGRD, '1' = 2, '2' = 1)
cwb_3_r <- recode(dat$EREPGRD, '1' = 2, '2' = 1)
cwb_3_r

# check that cwb_3 was reverse coded correctly 
table(cwb_3)
table(cwb_3_r)

cwb_4 <- (dat$ESPORT)

# mean scores for cwb items
cwb_1_mean <- mean(cwb_1)
cwb_2_r_mean <- mean(cwb_2_r)
cwb_3_r_mean <- mean(cwb_3_r)
cwb_4_mean <- mean (cwb_4)

# standard deviation for cwb items
cwb_1_sd <- sd(cwb_1)
cwb_2_r_sd <- sd(cwb_2_r)
cwb_3_r_sd <- sd(cwb_3_r)
cwb_4_sd <- sd(cwb_4)



# create new row of the cwb_scale based on the mean scores of all of the cwb variables
dat <- dat %>% dplyr::mutate(cwb_scale = rowMeans(pick(cwb_1, cwb_2, cwb_3, cwb_4)))
mean(dat$cwb_scale)
sd(dat$cwb_scale)

# assigning varibale names for residential stability
res_1 <- (dat$EEHC_PVTEN)
res_1
table(res_1)

res_2 <- (dat$EEHC_WHY)
res_2
table(res_2)

# assigning variable names for demographic data
dem_1 <- (dat$TAGE)
dem_1
dem_2 <- (dat$ERACE)
dem_2
dem_3 <- (dat$ESEX)
dem_3




```


# Methods
The current study was not preregistered. Data and code are available on [GitHub *click here* ](https://github.com/libby-kramer/Residential_Instability_Project)
Data retrieved from the 2024 Survey of Income and Program Participation <https://www.census.gov/programs-surveys/sipp/data/datasets/2024-data/2024.html>

## Participants
```{r demographics, include = FALSE}
# average age of respondents 
dem_1_mean <- mean(dem_1)
dem_1_mean

# standard deviation of age
dem_1_sd <- sd(dem_1)
dem_1_sd

```
The total sample in the current study consists of `r nrow(dat)`

## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

# Results
```{r descriptive table, echo=FALSE}
cwb_table <- dplyr::tribble(
    ~Item, ~Mean, ~SD,
    "CWB Item 1", cwb_1_mean,  cwb_1_sd,
    "CWB Item 2", cwb_2_r_mean,  cwb_2_r_sd,
    "CWB Item 3", cwb_3_r_mean,  cwb_3_r_sd,
    "CWB Item 4", cwb_4_mean,  cwb_4_sd)
cwb_table
```
# Discussion


\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
