---
title             : "Residential Instability and Childhood Wellbeing"
shorttitle        : "Residential Stability"

author: 
  - name          : "Libby Kramer"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Elliott Halll 75 E River Parkway, Minneapolis MN"
    email         : "krame937@umn.edu"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"

affiliation:
  - id            : "1"
    institution   : "University of Minnesota"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Thank you!

abstract: |
  *500 words*
  Research has suggested that quality of housing is a social determinant of health and that residential stability is a factor for older adults's health (CITE). This report will look more specifically at the relationship between residential stability and child well-being. As housing is not gaurenteed in the United States it is important to understand the relationship between where children are living and their wellbeing at home and in school. 
  In this study we found that...
  DO I NEED TO BE USING SIMULATED????

  
keywords          : "housing, residential stability, childhood wellbeing, health, social determinants of health"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library("groundhog")
library("dplyr")
library("psych")
library("labelled")
library("ggplot2")
library("data.table")
library("dplyr")
library("bit64")
install.packages("tidyverse")
library("tidyverse")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(0628)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

```{r data setup, include = FALSE}
# selecting variables from the full SIPP data set 
ds <- c("pu2024 2.csv")
pu <- fread(ds, sep = "|", select = c("ECLUB", "EEXPSCH", "EREPGRD", "ESPORT", "EEHC_PVTEN", "EEHC_WHY", "TAGE", "TDOB_BYEAR", "ERACE", "ESEX"))

names(pu) <- toupper(names(pu))

head(pu, 20)
dat <- na.omit(pu)
dat_sim <- read.csv("residential_stability_simulated_data_2025_11-17.csv")
dat_sim

# creating objects to be inserted into the report 
# assigning variables new names
cwb_1_sim <- (dat_sim$cwb_1_sim)



# recode cwb_2 to be keyed negatively
cwb_2_sim <- (dat_sim$cwb_2_sim)
recode(dat_sim$cwb_2_sim, '1'= 2, '2' = 1)
cwb_2_sim_r <- recode(dat_sim$cwb_2_sim, '1'= 2, '2' = 1)
cwb_2_sim_r

# checks that cwb_2 was reverse coded correctly
table(cwb_2_sim_r)
table(cwb_2_sim)

# recode cwb_3 to be keyed negatively
cwb_3_sim <- (dat_sim$cwb_3_sim)
recode(dat_sim$cwb_3_sim, '1' = 2, '2' = 1)
cwb_3_sim_r <- recode(dat_sim$cwb_3_sim, '1' = 2, '2' = 1)
cwb_3_sim_r

# check that cwb_3 was reverse coded correctly 
table(cwb_3_sim)
table(cwb_3_sim_r)

cwb_4 <- (dat$ESPORT)
cwb_4_sim <- (dat_sim$cwb_4_sim)
cwb_4_sim

# mean scores for cwb items
cwb_1_mean <- mean(cwb_1_sim)
cwb_2_r_mean <- mean(cwb_2_sim_r)
cwb_3_r_mean <- mean(cwb_3_sim_r)
cwb_4_mean <- mean (cwb_4_sim)

# standard deviation for cwb items
cwb_1_sd <- sd(cwb_1_sim)
cwb_2_r_sd <- sd(cwb_2_sim_r)
cwb_3_r_sd <- sd(cwb_3_sim_r)
cwb_4_sd <- sd(cwb_4_sim)


# create new row of the cwb_scale based on the mean scores of all of the cwb variables
### ERROR HERE
cwb_scale_sim <- (dat_sim$cwb_scale)

mean(dat$cwb_scale_sim)
sd(dat$cwb_scale_sim)

# assigning variable names for residential stability
res_1_sim <- (dat_sim$res_1_sim)
res_1_sim
table(res_1_sim)

res_2_sim <- (dat_sim$res_2_sim)
res_2_sim
table(res_2_sim)

# assigning variable names for demographic data
dem_1_sim <- (dat_sim$dem_1_sim)
dem_2_sim <- (dat_sim$dem_2_sim)
dem_3_sim <- (dat_sim$dem_3_sim)

```
## Present Study
  The purpose of the present study was to 

# Method
The current study was not preregistered. Data and code are available on Github [*click here* ](https://github.com/libby-kramer/Residential_Instability_Project)
Data retrieved from the 2024 Survey of Income and Program Participation <https://www.census.gov/programs-surveys/sipp/data/datasets/2024-data/2024.html>

## Participants and Procedure
```{r demographics, include = FALSE}
# average age of respondents 
dem_1_mean <- mean(dem_1_sim)
dem_1_mean

# standard deviation of age
dem_1_sd <- sd(dem_1_sim)
dem_1_sd
```

```{r demographic tables, include = FALSE}
summary(dem_2_sim)
dem_2_sim

race_percent <- results_dplyr <- dat_sim %>% 
  group_by(dem_2_sim) %>% 
  summarise(Percentage = n()/nrow(dat)*100)


apa_table(results_dplyr)

results_sex <- dat_sim %>% 
  group_by(dem_3_sim) %>% 
  summarise(Percentage = n()/nrow(dat)*100)
print(results_sex)
```

```{r tables, echo=FALSE}
apa_table(results_dplyr)
apa_table(results_sex)
```

FIX THIS WITH EMBEDDING VALUES
The sample size after omitting the missing values, due to the vast amount of missing, was 6691. The average age of the respondents was (*M* = `mean(dem_1))` with a standard deviation of `dem_1_sd`. The participants were 89.8% female, 77.7% white, 10.5% Black, 7.90% Asian, and 3.48% reported as residual. This sample size has some significant bias around sex and race. There answer options for the demographics were very binary.

## Measures

### Childhood well-being scale
### Residential Stability
## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

# Results
```{r descriptive table, echo=FALSE}
cwb_table <- dplyr::tribble(
    ~Item, ~Mean, ~SD,
    "CWB Item 1", cwb_1_mean,  cwb_1_sd,
    "CWB Item 2", cwb_2_r_mean,  cwb_2_r_sd,
    "CWB Item 3", cwb_3_r_mean,  cwb_3_r_sd,
    "CWB Item 4", cwb_4_mean,  cwb_4_sd)

apa_table(cwb_table, 
             caption = "Descriptivse for Childhood Wellbeing Scale Items",
             align = "l", digits = 2,
          note = "True or false questions keyed 1 as postive and 2 as negative")

```

cor.test(dat_sim$cwb_scale_sim, res_1_sim, method = c("pearson"))

```

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
